---
title: "Exploratory Data Analysis ~ R Coursera Project"
subtitle: "Data on Individual Household Electric Power Consumption"
author: "Shibani S Otis"
date: "8/21/2020"
output: html_document
---

```{r setup, echo=FALSE}
# The "echo = FALSE" option hides this chunk of code 
knitr::opts_knit$set(root.dir = "~/Documents/Coursera Data Science Specialization: Using R/Data Sets/")
```

### Introduction

This assignment uses data from the UC Irvine Machine Learning Repository, a popular repository for machine learning datasets. In particular, we will be using the "Household Electric Power Consumption" dataset which measures the electric power consumption in one household with a one-minute sampling rate over a period of almost 4 years. Different electrical quantities and some sub-metering values are available.

The following descriptions of the 9 variables in the dataset are taken from the UCI web site:

1. **Date**: Date in format dd/mm/yyyy
2. **Time**: time in format hh:mm:ss
3. **Global_active_power**: household global minute-averaged active power (in kilowatt)
4. **Global_reactive_power**: household global minute-averaged reactive power (in kilowatt)
5. **Voltage**: minute-averaged voltage (in volt)
6. **Global_intensity**: household global minute-averaged current intensity (in ampere)
7. **Sub_metering_1**: energy sub-metering No. 1 (in watt-hour of active energy). It corresponds to the kitchen, containing mainly a dishwasher, an oven and a microwave (hot plates are not electric but gas powered).
8. **Sub_metering_2**: energy sub-metering No. 2 (in watt-hour of active energy). It corresponds to the laundry room, containing a washing-machine, a tumble-drier, a refrigerator and a light.
9. **Sub_metering_3**: energy sub-metering No. 3 (in watt-hour of active energy). It corresponds to an electric water-heater and an air-conditioner.

The dataset has 2,075,259 rows and 9 columns. We will only be using data from the dates 2007-02-01 and 2007-02-02. 
    
### Data Processing

Let's load the electric power consumption data:

```{r}
# Note: Missing values are coded in as "?" in this data set
powerData <- read.table("household_power_consumption.txt", header = T, sep = ";", na.strings = "?", colClasses = c("character", "character", rep("numeric", 7)))

head(powerData)
```
&nbsp;  

Since we will only be using data from 2007-02-01 and 2007-02-02, we need to subset the data:

```{r}
# We need an "OR" operator i.e. | when using subset. Note, concatenate doesn't work here! 
subsetFeb <- subset(powerData, Date == "1/2/2007" | Date == "2/2/2007")

head(subsetFeb)
```

&nbsp; 

Now let's plot our first plot, a histogram and save it as a PNG file with a width of 480 pixels and a height of 480 pixels: 

```{r}
# First we need to specify where we want to save the plot  
#png(file="~/Documents/Coursera Data Science Specialization: Using R/Coursera Projects/Exploratory Data Analysis/plot1.png", width = 480, height = 480)

# Then we run the plot we want to save
hist(subsetFeb$Global_active_power, col = "red", xlab = "Global Active Power (kilowatts)", main = "Global Active Power")

# Lastly, turn the PNG graphics device off
#dev.off()        
```
&nbsp;  

Now we need to convert the "Date" and "Time" columns to the appropriate format:

```{r}
# Created a new column which took dates & times from the subset data, formatted into POSIXct
subsetFeb$DateTime <- as.POSIXct(paste(subsetFeb$Date, subsetFeb$Time), format = "%d/%m/%Y %H:%M:%S")

# Alternatively, we could have used the "strptime" function which uses POSIXlt:
#subsetFeb$DateTime <- strptime(paste(subsetFeb$Date, subsetFeb$Time), format = "%d/%m/%Y %H:%M:%S")

# Re-ordered the columns in the data set  
powerFinal <- subsetFeb[ ,c(1,2,10,3:9)]

head(powerFinal)
```
&nbsp; 

The second plot is a line graph. Let's save this plot as a 480 pixels by 480 pixels PNG:

```{r}
#png(file="~/Documents/Coursera Data Science Specialization: Using R/Coursera Projects/Exploratory Data Analysis/plot2.png", width = 480, height = 480)

plot(powerFinal$DateTime, powerFinal$Global_active_power, type = "l", xlab = "", ylab = "Global Active Power (kilowatt)")
#dev.off()
```

&nbsp; 

Now to make the third plot and save as a PNG:

```{r}
#png(file="~/Documents/Coursera Data Science Specialization: Using R/Coursera Projects/Exploratory Data Analysis/Project 1 Plots/plot3.png", width = 480, height = 480)

plot(powerFinal$DateTime, powerFinal$Sub_metering_1, type = "l", xlab = "", ylab = "Energy sub metering")

# We need to add additional lines to the plot, thus we use the "lines" function instead of "plot"
lines(powerFinal$DateTime, powerFinal$Sub_metering_2, type = "l", col = "red", xlab = "", ylab = "Energy sub metering")
lines(powerFinal$DateTime, powerFinal$Sub_metering_3, type = "l", col = "blue", xlab = "", ylab = "Energy sub metering")

# Created a legend
legend("topright", lty = 1, col = c("black", "red", "blue"), legend = c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"))
#dev.off()
```

&nbsp; 

The last plot consists of the 2nd and 3rd plots, along with two other plots saved as a PNG file:

```{r}
#png(file="~/Documents/Coursera Data Science Specialization: Using R/Coursera Projects/Exploratory Data Analysis/plot4.png", width = 480, height = 480)

# We need 2 rows and 2 columns for our plots and an appropriate margin to fit them all
par(mfcol = c(2,2), mar = c(5,5,2,2))

plot(powerFinal$DateTime, powerFinal$Global_active_power, type = "l", xlab = "", ylab = "Global Active Power")

plot(powerFinal$DateTime, powerFinal$Sub_metering_1, type = "l", xlab = "", ylab = "Energy sub metering")
lines(powerFinal$DateTime, powerFinal$Sub_metering_2, type = "l", col = "red", xlab = "", ylab = "Energy sub metering")
lines(powerFinal$DateTime, powerFinal$Sub_metering_3, type = "l", col = "blue", xlab = "", ylab = "Energy sub metering")
legend("topright", lty = 1, col = c("black", "red", "blue"), legend = c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"), cex = 0.75, bty = "n")   # bty ="n" removes box around the legend; cex reduces its size 

plot(powerFinal$DateTime, powerFinal$Voltage, type = "l", xlab = "datetime", ylab = "Voltage")

plot(powerFinal$DateTime, powerFinal$Global_reactive_power, type = "l", xlab = "datetime", ylab = "Global_reactive_power")
#dev.off()
```

